---
title: ""
author: ""
date: "23 tammikuuta 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Install packages if needed
# Load libraries
library(car)
library(Epi)
library(epitools)
library(ggplot2)
library(Hmisc)
library(knitr)
library(mfp)
library(mgcv)
library(muhaz)
library(rstpm2)
library(survival)
library(survminer)
library(PerformanceAnalytics)
library(scales)
library(xtable)
library(tidyr)
library(genoPlotR)

```

# Figure 1

Distribution of somatic insertions across 202 colorectal tumors

```{r, Figure 1a}

# Open file
Data<-read.table("Clinical.txt",sep="\t",header=T,stringsAsFactors=F,row.names=1) #Clinical.txt -> Supplementary Dataset 5; Insertion count = ins

## Plot
h1 <- hist(Data$ins,xlab="Insertion count", ylab="Frequency", col="darkslategrey",breaks = 20, axes=FALSE, main="",cex.lab=1.5)
axis (side=1, at=seq(0,205,by=10))
axis(side=2, at= seq(0,80,by=10))
title("a)", adj=0)

## Export plot

svg(filename="Figure_1a.svg")
h1 <- hist(Data$ins,xlab="Insertion count", ylab="Frequency", col="darkslategrey",breaks = 20, axes=FALSE, main="",cex.lab=1.5)
axis (side=1, at=seq(0,205,by=10))
axis(side=2, at= seq(0,80,by=10))
title("a)", adj=0)
dev.off()

```

Insertion density over replication time. The genome was stratified by replication time in five categories where 0 referred to the earliest replication timing. Replication time was extracted from Chen, C.-L. et al. Impact of replication timing on non-CpG and CpG substitution rates in mammalian genomes. Genome Res. 20, 447–457 (2010).

```{r, Figure 1b}

# Open file
Density <-read.table("Density.txt",sep="\t",header=T,stringsAsFactors=FALSE,row.names=NULL )

# Plot
boxplot(Density$early.d,Density$mid.early.d,Density$mid.d,Density$mid.late.d,Density$late.d,names=c("0-0.2","0.2-0.4","0.4-0.6","0.6-0.8","0.8-1"),outline=F, cex.axis=0.90, ylim= c(0,1e-07), xlab="Replication time fraction", ylab="Insertion density per tumor/bp", cex.lab=1.5)
title("b)", adj=0)

stripchart(list(Density$early.d,Density$mid.early.d,Density$mid.d,Density$mid.late.d,Density$late.d), vertical=T, add=T, method="jitter",
pch=20,cex=1,col=(adjustcolor("darkslategrey",alpha.f=0.3)))

# Export plot

svg(filename="Figure_1b.svg")
boxplot(Density$early.d,Density$mid.early.d,Density$mid.d,Density$mid.late.d,Density$late.d,names=c("0-0.2","0.2-0.4","0.4-0.6","0.6-0.8","0.8-1"),outline=F, cex.axis=0.90, ylim= c(0,1e-07), xlab="Replication time fraction", ylab="Insertion density per tumor/bp", cex.lab=1.5)
title("b)", adj=0)

stripchart(list(Density$early.d,Density$mid.early.d,Density$mid.d,Density$mid.late.d,Density$late.d), vertical=T, add=T, method="jitter",
pch=20,cex=1,col=(adjustcolor("darkslategrey",alpha.f=0.3)))

dev.off()

```

# Figure 2

Retrotransposon insertions in protein-coding genes (GRCh37_87).Gene expression values have been deposited in the Zenodo database under the Digital Object Identifier [10.5281/zenodo.3241399] [https://zenodo.org/deposit/3241399]. 

```{r Figure 2}

## Set color
colors1 <- "darkblue"

## Open Data
Genes <-read.table("Genes.txt",sep="\t",header=T,stringsAsFactors=FALSE,row.names=NULL) 
Genes$ins.cat <- cut(Genes$gene,breaks=c(0,1,2,3,4,Inf),right=F)

## Set color
colors1 <- adjustcolor(colors1, alpha.f = 0.3)

# Plot
box.exp <- ggplot(data=Genes, aes(x= ins.cat,y= log1p(tpm)), outlier=F)
box.exp + geom_boxplot(fill= "white", outlier.shape=NA) + labs(x= "Gene insertion count",y="Log (median tpm)") + geom_point(position = position_jitter(w=0.3), col= colors1, size=0.5)+ theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + scale_x_discrete(labels=c("0", "1", "2","3","> 4"))+ labs(title="")


# Export plot

svg(filename="Figure_2.svg")
box.exp <- ggplot(data=Genes, aes(x= ins.cat,y= log1p(tpm)), outlier=F)
box.exp + geom_boxplot(fill= "white", outlier.shape=NA) + labs(x= "Gene insertion count",y="Log (median tpm)") + geom_point(position = position_jitter(w=0.3), col= colors1, size=0.5)+ theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + scale_x_discrete(labels=c("0", "1", "2","3","> 4"))+ labs(title="")
dev.off()

```

# Figure 3

Retrotransposon insertion distribution in LRP1B and DLG2.

```{r}

setwd("P:/h304/aaltonen/Tatiana/PhD/WGS/Retro_CRC/Somatic/Manuscript_files/Submission_NatComms/NatComms_revision/Top_genes/GenoplotR")

lrp1b <-read.table("LRP1B_270119.txt",sep="\t",header=T,stringsAsFactors=FALSE,row.names=NULL) # Instead Source Data file 1 can be used
dlg2 <-read.table("DLG2_270119.txt",sep="\t",header=T,stringsAsFactors=FALSE,row.names=NULL) # Instead Source Data file 2 can be used
gene_seg_lrp1b <- dna_seg(lrp1b)
gene_seg_dlg2 <- dna_seg(dlg2)

# Plot gene map

plot_gene_map(list(gene_seg_lrp1b,gene_seg_dlg2),dna_seg_scale= TRUE)

# Plot gene map


svg(filename="LRP1B_270119.svg", width=17/2.54, height=10/2.54)
plot_gene_map(list(gene_seg_lrp1b),dna_seg_scale= TRUE)
dev.off()


svg(filename="DLG2_270119.svg", width=17/2.54, height=10/2.54)
plot_gene_map(list(gene_seg_dlg2),dna_seg_scale= TRUE)
dev.off()

```

# Figure 5 
Insertion and AI frequency in 21 fragile sites. a) Insertion fraction over the fraction of allelic imbalance in 21 fragile sites. b) Gene expression (median TPM values from 34 tumors) in fragile sites with high insertion fraction and fragile sites with high allelic imbalance fraction. Fragilesites.txt corresponds to Supplementary Dataset 3 with the following variable name conversion: 

Insertion fraction (n=5072) = ins.fr
Focal AI events fraction (n=105527) = ai.fr
Category = cat
Gene expression (Median TPM) = tpm

```{r Figure 5a}
# Open data
FS <-read.table("FragileSites.txt",sep="\t",header=T,stringsAsFactors=FALSE,row.names=1) # Supplementary Dataset 3
   
# Plot
Fig.4a <- ggplot(data=FS,aes(x =ai.fr,y=ins.fr, color=cat))
Fig.4a <- Fig.4a + geom_point(pch=20,cex=10,col=adjustcolor("dodgerblue4",alpha.f=0.6)) + labs(title="a)",x="Allelic imbalance fraction", y="Insertion fraction") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(),axis.line = element_line(colour = "black"),title = element_text(vjust=2), axis.text = element_text(size=12), axis.title = element_text(size=15)) 

Fig.4a
# Export plot
svg(filename="Figure_5a.svg")
Fig.4a <- ggplot(data=FS,aes(x =ai.fr,y=ins.fr, color=cat))
Fig.4a <- Fig.4a + geom_point(pch=20,cex=10,col=adjustcolor("dodgerblue4",alpha.f=0.6)) + labs(title="a)",x="Allelic imbalance fraction", y="Insertion fraction") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(),axis.line = element_line(colour = "black"),title = element_text(vjust=2), axis.text = element_text(size=12), axis.title = element_text(size=15)) 
dev.off()

```



```{r Figure 5b}

# Plot    
Fig.4b <- ggplot(na.omit(FS), aes(x=as.factor(cat),y=tpm))
Fig.4b + geom_boxplot(outlier.shape=NA) + labs(title="b)",x= "fragile site classification",y="Colorectal tumors median tpm") + geom_point(position = position_jitter(w=0.3),col=adjustcolor("dodgerblue4",alpha.f=0.6),size=7)+ theme_bw() + 
theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),title = element_text(vjust=2), axis.text = element_text(size=12), axis.title = element_text(size=15))


# Export plot
svg(filename="Figure_5b.svg")
Fig.4b <- ggplot(na.omit(FS), aes(x=as.factor(cat),y=tpm))
Fig.4b + geom_boxplot(outlier.shape=NA) + labs(title="b)",x= "Fragile site classification",y="Colorectal tumors median tpm") + geom_point(position = position_jitter(w=0.3),col=adjustcolor("dodgerblue4",alpha.f=0.6),size=7)+ theme_bw() + 
theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),title = element_text(vjust=2), axis.text = element_text(size=12), axis.title = element_text(size=15))
dev.off()

```


# Multiple Linear model

Supplementary Dataset 5 can be used instead of clinical.txt by changing the variable names: 

* Insertion count  = ins
* CIMP = cimp
* Allelic imbalance (bp) = aibp
* BRAF = braf
* TP53 = tp53
* Mean coverage = cov.mean
* RER = rer
* Dukes tumor stage = dukes
* Age at diagnosis = age
* Sex = sex
* Tumor location = location
* Times (days) = days
* Event = status.crc


```{r Linear model}

# Open data
Data<-read.table("Clinical.txt",sep="\t",header=T,stringsAsFactors=F,row.names=1) # clinical.txt = Supplementary Dataset 5
Data$ins.cat<-cut(Data$ins,breaks=c(0,20,Inf),right=F)
Data$age.cat<-cut(Data$age,breaks=c(0,55,75,Inf),right=F)
Data$rer<-as.factor(Data$rer);Data$rer<-relevel(Data$rer,ref="MSS") # might not be needed 
Data.NM<-Data[,c("ins","ins.cat","age","age.cat","sex","dukes","location","rer","aibp","cimp","braf","days","status.crc","status.os")] # For cox proportional hazards model; status.os not used
Data.NM<-Data.NM[complete.cases(Data.NM),] # For cox proportional hazards model


# Linear model

fit.lm<-lm(log1p(ins)~cimp+I(aibp/(3*10^8))+tp53+rer+I(cov.mean/10)+I(age/10)+sex+dukes+location,data=Data)
summary(fit.lm)

# Linear model without influential sample 
fit.lm.noinf <- lm(log1p(ins)~cimp+I(aibp/(3*10^8))+tp53+rer+I(cov.mean/10)+I(age/10)+sex+dukes+location,data=Data[!row.names(Data)%in%"c439_1",])
summary(fit.lm.noinf)


# Linear model including BRAF mut

fit.lm.braf<-lm(log1p(ins)~cimp+I(aibp/(3*10^8))+tp53+rer+I(cov.mean/10)+I(age/10)+sex+dukes+location+braf,data=Data)
summary(fit.lm.braf)


# Linear model only including MSS

Data.MSS <- subset(Data, rer=="MSS")

fit.lm.MSS <- lm(log1p(ins)~cimp+I(aibp/(3*10^8))+tp53+I(cov.mean/10)+I(age/10)+sex+dukes+location,data=Data.MSS)
summary(fit.lm.MSS)

```


```{r Figure S3,S4,S5}

# Model diagnostics

par(mfrow=c(2,3))
plot(fit.lm,which=1:6)


# Export Figure S3

svg(filename="figure_S3.svg")
par(mfrow=c(2,3))
plot(fit.lm,which=1:6)
dev.off()

# Residuals categorical variables

for(i in c("cimp","rer","dukes","sex","location","tp53")){
  boxplot(residuals(fit.lm)~fit.lm$model[,i],
    ylab="residuals",
    xlab=i,
    outline=F,
    ylim=range(residuals(fit.lm))
    )
  stripchart(residuals(fit.lm)~fit.lm$model[,i],
    vertical=T,
    method="jitter",
    pch=1,
    add=T,
    data=Data)
  abline(h=0)
}

# Export  figure S4

svg(filename = "Figure_S4.svg")
for(i in c("cimp","rer","dukes","sex","location","tp53")){
  boxplot(residuals(fit.lm)~fit.lm$model[,i],
    ylab="residuals",
    xlab=i,
    outline=F,
    ylim=range(residuals(fit.lm))
    )
  stripchart(residuals(fit.lm)~fit.lm$model[,i],
    vertical=T,
    method="jitter",
    pch=1,
    add=T,
    data=Data)
  abline(h=0)
}

dev.off()

# Residuals continous variables

par(mfrow=c(2,2))
for(i in c("I(age/10)","I(cov.mean/10)","I(aibp/(3 * 10^8))")){
  scatter.smooth(fit.lm$model[,i],residuals(fit.lm),
    xlab=i)
  abline(h=0)
}
par(mfrow=c(1,1))

# Export Figure_S5

svg(filename="Figure_S5")
par(mfrow=c(2,2))
for(i in c("I(age/10)","I(cov.mean/10)","I(aibp/(3 * 10^8))")){
  scatter.smooth(fit.lm$model[,i],residuals(fit.lm),
    xlab=i)
  abline(h=0)
}
par(mfrow=c(1,1))


dev.off()

#Goodness-of-fit

pchisq(sum(residuals(fit.lm)**2),fit.lm$df.residual,lower.tail=F)

#Collinearity

vif(fit.lm)

#Functional form

par(mfrow=c(2,2))
plot(gam(log1p(ins)~cimp+s(aibp)+tp53+rer+s(cov.mean)+s(age)+sex+dukes+location,
  data=Data)
)
par(mfrow=c(1,1))
mfp(log1p(ins)~cimp+fp(aibp)+tp53+rer+fp(cov.mean)+fp(age)+sex+dukes+location,
  alpha=0.15,
  verbose=T,
  data=Data
)
```


#  Cox Proportionals Hazards model

```{r, Survival}

# survival object

surv.dss<-Surv(Data.NM$days,Data.NM$status.crc)

# Kaplan-Meier estimates

survfit.ins.cat<-survfit(surv.dss~ins.cat,data=Data.NM)
names(survfit.ins.cat$strata)<-levels(Data.NM$ins.cat)

#Cox model

fit.coxph.dss<-coxph(Surv(days,status.crc)~I(ins/10)+rer+cimp+braf+age.cat+sex+dukes+strata(location)+I(aibp/(3*10^8)),data=Data.NM)

#Grambsch-Therneau test

cox.zph(fit.coxph.dss)

#Scaled Schoenfeld residuals


ggcoxdiagnostics(fit.coxph.dss,type="scaledsch",ox.scale="time")


# Export figure S6

svg(filename="Figure_S6.svg")
ggcoxdiagnostics(fit.coxph.dss,type="scaledsch",ox.scale="time")
dev.off()

#DfBeta

ggcoxdiagnostics(fit.coxph.dss, type = "dfbeta",
                 linear.predictions = FALSE, ggtheme = theme_bw())

# Export Figure S7 

svg(filename="Figure_S7.svg")
ggcoxdiagnostics(fit.coxph.dss, type = "dfbeta",
                 linear.predictions = FALSE, ggtheme = theme_bw())
dev.off()

#Martingale residuals vs. likelihood displacement

ld<-numeric()
for(i in 1:fit.coxph.dss$n){
  dB<-residuals(fit.coxph.dss,type="dfbeta")[i,]
  ld[i]<-t(dB)%*%solve(fit.coxph.dss$var)%*%dB
}
plot(residuals(fit.coxph.dss,type="martingale"),ld,pch=3,xlab="Martingale residual",ylab="Likelihood displacement")


# Export Figure S8

svg(filename="Figure_S8.svg")
ld<-numeric()
for(i in 1:fit.coxph.dss$n){
  dB<-residuals(fit.coxph.dss,type="dfbeta")[i,]
  ld[i]<-t(dB)%*%solve(fit.coxph.dss$var)%*%dB
}
plot(residuals(fit.coxph.dss,type="martingale"),ld,pch=3,xlab="Martingale residual",ylab="Likelihood displacement")
dev.off()


```



```{r Kaplan-Meier curves 1,fig.width=7,fig.height=7,fig.cap="Kaplan-Meier curves by insertion count.",out.width="60%",warning=FALSE}
  ggsurvplot(survfit.ins.cat,
  size=0.5,
  xscale="d_y",
  xlim=c(0,15*365.25),
  break.x.by=2.5*365.25,
  risk.table=T,
  risk.table.col="strata",
  pval=F,
  pval.coord=c(12.5*365.25,0.9),
  xlab="Time after operation (years)",
  ylab="Disease-specific survival",
  font.title=c("14","bold","black"),
  font.legend=c(13,"plain","black"),
  palette=palette()[-1]
)



#Colors
palette(c("black","#00468B","#BC3C29FF","#0099B4","#868686","#EFC000FF","#20854E","#00468B99","#BC3C2999","#0099B499","#86868699","#20854E99","#E1872799"))


# Export Figure 6 

svg(filename ="Figure_6.svg")
  ggsurvplot(survfit.ins.cat,
  size=0.5,
  xscale="d_y",
  xlim=c(0,15*365.25),
  break.x.by=2.5*365.25,
  risk.table=T,
  risk.table.col="strata",
  pval=F,
  pval.coord=c(12.5*365.25,0.9),
  xlab="Time after operation (years)",
  ylab="Disease-specific survival",
  font.title=c("14","bold","black"),
  font.legend=c(13,"plain","black"),
  palette=palette()[-1]
)
  dev.off()


```
